
---
Assignment 2




```{r}
library(readxl)
```

```{r}
temp = tempfile(fileext = ".xlsx")
```

```{r}
dataURL <- "https://raw.githubusercontent.com/popester007/data/master/project%20data%20GHI.csv"
```

```{r}
download.file(dataURL, destfile =temp, mode='wb')
```

```{r}
ghIndex = read.csv(dataURL)
```

```{r}
head(ghIndex,51)
```

```{r}
str(ghIndex)
```


```{r}
names(ghIndex)
```

```{r}
summary(ghIndex)
```

```{r}
names(ghIndex) <- c("Rank", "Country", "1992", "2000", "2008", "2017")
```

```{r}
names(ghIndex)
```

```{r}
library(car)
scatterplotMatrix(ghIndex[,c(3:6)], diagonal='histogram',reg.line=NULL, smoother=NULL)
```

```{r}
str(ghIndex)
```

```{r}
summary(ghIndex)
```

Pick one variable to explore: 1992
```{r}
summary(ghIndex$'1992',na.rm = T)
```
Analyze dispersion
```{r}
sd(ghIndex$'1992',na.rm = T)
```

```{r}
library(e1071)
```


Analyze Skewness
```{r}
skewness(ghIndex$'1992',na.rm = T)
```

Analyze Kurtosis
```{r}
kurtosis(ghIndex$'1992',na.rm = T)
```
Next step ?
```{r}
data=ghIndex[is.finite(ghIndex$'1992'),]
```
selecting a variable
```{r}
var=data$'1992'
```
saving mean and sd
```{r}
mnVar=mean(var,na.rm = T)
sdVar=sd(var,na.rm = T)
```
plotting
```{r}
library(ggplot2)
base = ggplot(data, aes(x=var))
hist = base + geom_histogram(fill="green",
                             color='grey',
                             aes(y=..density..))
```
next step...?
```{r}
histAndNormal = hist + stat_function(fun=dnorm,
                                     color="red",
                                     args=list(mean=mnVar,sd=sdVar))
```

```{r}
histAndNormal
```

now histogram with central measures
```{r}
mdVar=median(var)
```

now histogram with central measures
```{r}
histAndNormal + geom_vline(aes(xintercept = mnVar, colour="mean"),
                           show.legend = TRUE) +
  geom_vline(aes(xintercept = mdVar,colour="median"),
             show.ledgend = TRUE) +
  scale_color_manual(name = "centrals",
                     values = c(median = "blue", mean = "red"))
```
Now for outliers
```{r}
ghIndex$'1992'=as.numeric(ghIndex$'1992')
```

```{r}
summary(ghIndex$'1992')
```
looking at 25th and 75th percentile
```{r}
(bp=boxplot(ghIndex$'1992',horizontal = T))
```

```{r}
bp$stats
```

compute IQR
```{r}
(iqr=IQR(ghIndex$'1992',na.rm = T))
```
For 75th Q:
```{r}
q75=bp$stats[4]
(capHigh = q75 + iqr*1.5)
```

For 25th Q:
```{r}
q25=bp$stats[2]
(capLow=q25 - iqr*1.5)
```

Any value above the High value (73.575) or below the Low value (15.775) is an outlier

```{r}
length(bp$out)
```

```{r}
StdDev=sd(ghIndex$'1992',na.rm = T)
Mean=mean(ghIndex$'1992',na.rm = T)
```

```{r}
(lowCapT=Mean-2*StdDev)
```

```{r}
(upCapT=Mean+2*StdDev)
```

making maps
```{r}
compressedMap= 'https://github.com/popester007/PROJECT/blob/master/TM_WORLD_BORDERS_SIMPL-0.3.zip'
temp=tempfile()
download.file(compressedMap, temp)
unzip(temp)
```


