```{r}
library(readxl)
```

```{r}
temp = tempfile(fileext = ".xlsx")
```

```{r}
dataURL <- "https://raw.githubusercontent.com/popester007/data/master/project%20data%20GHI.csv"
```

```{r}
download.file(dataURL, destfile =temp, mode='wb')
```

```{r}
ghIndex = read.csv(dataURL)
```

```{r}
head(ghIndex,51)
```

```{r}
str(ghIndex)
```

```{r}
names(ghIndex)
```

```{r}
summary(ghIndex)
```

```{r}
names(ghIndex) <- c("Rank", "Country", "1992", "2000", "2008", "2017")
```

```{r}
names(ghIndex)
```

```{r}
library(car)
scatterplotMatrix(ghIndex[,c(3:6)], diagonal='histogram',reg.line=NULL, smoother=NULL)
```

```{r}
str(ghIndex)
```

```{r}
summary(ghIndex)
```

Univariate analysis
```{r}
summary(ghIndex$'1992',na.rm = T)
```

Analyze dispersion
```{r}
sd(ghIndex$'1992',na.rm = T)
```

```{r}
library(e1071)
```

Analyze skewness
```{r}
skewness(ghIndex$'1992',na.rm = T)
```

Analyze Kurtosis
```{r}
kurtosis(ghIndex$'1992',na.rm = T)
```

```{r}
data=ghIndex[is.finite(ghIndex$'1992'),]
```

Select variable
```{r}
var=data$'1992'
```

Saving mean and SD
```{r}
mnVar=mean(var,na.rm = T)
sdVar=sd(var,na.rm = T)
```

Plotting
```{r}
library(ggplot2)
base = ggplot(data, aes(x=var))
hist = base + geom_histogram(fill="green",
                             color='grey',
                             aes(y=..density..))
```

```{r}
histAndNormal = hist + stat_function(fun=dnorm,
                                     color="red",
                                     args=list(mean=mnVar,sd=sdVar))
```

```{r}
histAndNormal
```

now histogram with central measures
```{r}
mdVar=median(var)
```

```{r}
histAndNormal + geom_vline(aes(xintercept = mnVar, colour="mean"),
                           show.legend = TRUE) +
  geom_vline(aes(xintercept = mdVar,colour="median"),
             show.ledgend = TRUE) +
  scale_color_manual(name = "centrals",
                     values = c(median = "blue", mean = "red"))
```

Now for outliers
```{r}
ghIndex$'1992'=as.numeric(ghIndex$'1992')
```

```{r}
summary(ghIndex$'1992')
```

Looking at 27th and 75th percentile
```{r}
(bp=boxplot(ghIndex$'1992',horizontal = T))
```

```{r}
bp$stats
```

Calculating IQR
```{r}
(iqr=IQR(ghIndex$'1992',na.rm = T))
```

75th Q
```{r}
q75=bp$stats[4]
(capHigh = q75 + iqr*1.5)
```

For 25th Q
```{r}
q25=bp$stats[2]
(capLow=q25 - iqr*1.5)
```

Any value above the High value (73.575) or below the Low value (15.775) is an outlier
```{r}
length(bp$out)
```

```{r}
StdDev=sd(ghIndex$'1992',na.rm = T)
Mean=mean(ghIndex$'1992',na.rm = T)
```

```{r}
(lowCapT=Mean-2*StdDev)
```

```{r}
(upCapT=Mean+2*StdDev)
```

making maps
```{r}
compressedMap= 'https://github.com/popester007/PROJECT/blob/master/TM_WORLD_BORDERS_SIMPL-0.3.zip?raw=true'
temp=tempfile()
download.file(compressedMap, temp)
unzip(temp)
```

```{r}
(maps=list.files(pattern = 'shp'))
```

Select Map
```{r}
library(rgdal)
worldMap <- rgdal::readOGR('TM_WORLD_BORDERS_SIMPL-0.3.shp',stringsAsFactors = F)
```

```{r}
names(worldMap)
```

```{r}
names(ghIndex)
```


```{r}
worldMapGHI=merge(worldMap,ghIndex, by.x='NAME', by.y='Country',all.x=F)
```

```{r}
plot(worldMap,col='black')
```
```{r}
worldMapGHI$`2017`
```

```{r}
library(RColorBrewer)
library(classInt)
```


Making a map to see 2017 data

```{r}
varToPlot2=worldMapGHI$`2017`
```

```{r}
numberOfClasses = 5
colorForScale='OrRd'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPlot2, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)
```

```{r}
legendText="GHI Index"
shrinkLegend=0.8
title="Global Hunger Index in 2017"

plot(worldMap,col='grey',main=title,border=200)
plot(worldMapGHI, col = colorPallette,border=200,add=T)

legend('left', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```

Making a map to see 2008 data
```{r}
varToPlot3=worldMapGHI$`2008`
```

```{r}
numberOfClasses = 5
colorForScale='OrRd'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPlot3, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)
```

```{r}
legendText="GHI Index"
shrinkLegend=0.8
title="Global Hunger Index in 2008"

plot(worldMap,col='grey',main=title, border=200)
plot(worldMapGHI, col = colorPallette,border=200,add=T)

legend('left', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```

Making a map for 2000 data
```{r}
varToPlot4=worldMapGHI$`2000`
```

```{r}
numberOfClasses = 5
colorForScale='OrRd'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPlot4, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)
```

```{r}
legendText="GHI Index"
shrinkLegend=0.8
title="Global Hunger Index in 2000"

plot(worldMap,col='grey',main=title,border=200)
plot(worldMapGHI, col = colorPallette,border=200,add=T)

legend('left', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```


Making a map for 1992 data

```{r}
varToPlot5=worldMapGHI$`2000`
```

```{r}
numberOfClasses = 5
colorForScale='OrRd'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPlot5, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)
```

```{r}
legendText="GHI Index"
shrinkLegend=0.8
title="Global Hunger Index in 2000"

plot(worldMap,col='grey',main=title,border=200)
plot(worldMapGHI, col = colorPallette,border=200,add=T)

legend('left', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```

Make a new variable to calculate the difference between 2017 and 2008, and then map it
```{r}

```

